---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(tibble.print_min = 5L, tibble.print_max = 5L)
```

# ifo

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/m-muecke/ifo/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/m-muecke/ifo/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/ifo)](https://CRAN.R-project.org/package=ifo)
<!-- badges: end -->

## Overview

The goal of ifo is to provide a simple interface to the [ifo institute](https://www.ifo.de/en/ifo-time-series) time series data.
The package is still in an early stage of development and the API might change in the future.
Feel free to open an issue if you have any suggestions.

## Installation

You can install the released version of ifo from [CRAN](https://CRAN.R-project.org) with:

```{r, eval = FALSE}
install.packages("ifo")
```

And the development version from [GitHub](https://github.com/) with:

```{r, eval = FALSE}
# install.packages("pak")
pak::pak("m-muecke/ifo")
```

## Usage

```{r demo, message = FALSE}
library(ifo)

climate <- ifo_business()
head(climate)
```

```{r plotting, message = FALSE, echo = FALSE, dpi = 300}
library(ggplot2)

climate |>
  subset(series == "index" & yearmonth >= "2015-01-01") |>
  ggplot(aes(x = yearmonth, y = value, color = indicator)) +
  geom_line() +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    legend.position = "top",
    plot.title = element_text(face = "bold"),
    plot.caption = element_text(
      hjust = 0, vjust = 0, size = 8, margin = margin(10, 0, 0, 0)
    ),
    panel.grid.major.y = element_line(color = "black", linewidth = 0.2),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(color = "black"),
    axis.title = element_blank(),
    plot.margin = margin(10, 10, 10, 10)
  ) +
  labs(
    title = "ifo Business Climate Germany", subtitle = "Seasonally adjusted",
    caption = sprintf(
      "Source: ifo Business Survey, %s.", format(max(climate$yearmonth), "%B %Y")
    )
  ) +
  scale_color_manual(
    values = c(climate = "darkred", situation = "darkgrey", expectation = "darkblue"),
    labels = c(
      "ifo Business Climate",
      "Assessment of business situtation",
      "Business expectation"
    )
  )
```
