---
title: "ifo Business Climate Index"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ifo Publication}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The following will demonstrate how to use the `ifo` package to access the ifo Business Climate data
and reproduce some of the plots from the ifo website.

```{r theme, message = FALSE}
library(ggplot2)

ifo_theme <- theme_minimal() +
  theme(
    legend.title = element_blank(),
    legend.position = "top",
    plot.title = element_text(face = "bold"),
    plot.caption = element_text(
      hjust = 0, vjust = 0, size = 8, margin = margin(10, 0, 0, 0)
    ),
    panel.grid.major.y = element_line(color = "black", linewidth = 0.2),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(color = "black"),
    plot.margin = margin(10, 10, 10, 10)
  )
```

### ifo Business Climate Germany

```{r climate, message = FALSE, dpi = 300}
library(dplyr)
library(ifo)

climate <- ifo_climate()

climate |>
  select(yearmonth, ends_with("index")) |>
  tidyr::pivot_longer(-yearmonth, names_to = "component") |>
  mutate(component = sub("_index", "", component, fixed = TRUE)) |>
  ggplot(aes(x = yearmonth, y = value, color = component)) +
  geom_line() +
  labs(
    title = "ifo Business Climate Germany", subtitle = "Seasonally adjusted",
    y = NULL, x = NULL,
    caption = sprintf(
      "Source: ifo Business Survey, %s.", format(max(climate$yearmonth), "%B %Y")
    )
  ) +
  scale_color_manual(
    values = c(climate = "darkred", situation = "darkgrey", expectation = "darkblue"),
    labels = c(
      "ifo Business Climate",
      "Assessment of business situtation",
      "Business expectation"
    )
  ) +
  ifo_theme
```

### ifo Business Climate by Sector

```{r sector, dpi = 300}
sectors <- ifo_climate("sectors")

sectors |>
  select(yearmonth, ends_with("balance")) |>
  tidyr::pivot_longer(-yearmonth,
    names_to = c("component", "sector"), names_pattern = "(.*)_(.*)_balance"
  ) |>
  filter(
    sector %in% c("manufacturing", "services", "trade", "construction"),
    yearmonth >= "2020-01-01"
  ) |>
  mutate(
    sector = factor(sector,
      levels = c("manufacturing", "services", "trade", "construction")
    )
  ) |>
  ggplot(aes(x = yearmonth, y = value, color = component)) +
  geom_line() +
  facet_wrap(~sector,
    labeller = labeller(sector = c(
      manufacturing = "Manufacturing",
      services = "Service Sector",
      trade = "Trade",
      construction = "Construction"
    ))
  ) +
  labs(
    title = "ifo Business Climate, Business Situation, and Expectations by Sector",
    subtitle = "Balances, seasonally adjusted",
    y = NULL, x = NULL,
    caption = sprintf(
      "Source: ifo Business Survey, %s.", format(max(sectors$yearmonth), "%B %Y")
    )
  ) +
  scale_color_manual(
    values = c(climate = "darkred", situation = "darkgrey", expectation = "darkblue"),
    labels = c(
      "ifo Business Climate",
      "Assessment of business situtation",
      "Business expectation"
    )
  ) +
  ifo_theme
```

### ifo Business Uncertainty Germany

"The ifo Business Uncertainty measures how difficult it is for managers to predict the development of their company's business situation over the next six months. The measure is calculated based on the weighted fractions of companies that fall into the answer ..."

```{r uncertainty, dpi = 300}
climate |>
  select(yearmonth, uncertainty) |>
  tidyr::drop_na() |>
  ggplot(aes(x = yearmonth, y = uncertainty)) +
  geom_line() +
  labs(
    title = "ifo Business Uncertainty Germany", subtitle = "Original values",
    y = NULL, x = NULL,
    caption = sprintf(
      "Source: ifo Business Survey, %s.", format(max(climate$yearmonth), "%B %Y")
    )
  ) +
  ifo_theme
```
