---
title: "The ifo index as a leading indicator"
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  dpi = 300L
)
```

The ifo Business Climate Index is widely used as a *leading indicator* for the German
economy: because firms are surveyed about their current situation and their expectations
before the official statistics are compiled, the index is meant to signal turning points in
economic activity ahead of the hard data.
This article puts that claim to the test by adapting the evaluation framework of
[Schröder and Hüfner (2002)](https://ftp.zew.de/pub/zew-docs/dp/dp0256.pdf), who study how
useful sentiment indicators are for forecasting German economic activity.
[Lehmann (2023)](https://link.springer.com/article/10.1007/s41549-022-00079-5) provides a
broader review of the forecasting evidence from the ifo Business Survey.

Following [Fritsche and Stephan (2002)](https://www.diw.de/documents/publikationen/73/diw_01.c.38193.de/dp241.pdf),
we ask whether the ifo index satisfies three criteria of a good leading indicator:

1. **Co-movement** with the reference series (industrial production).
2. **One-directional causality** — the indicator should help predict activity, not merely
   react to it.
3. **Forecast value** — it should improve a forecast beyond what the activity series predicts
   about itself.

We use the ifo Business Climate together with the German industrial production index, which is
monthly (like the ifo survey) and therefore needs no mixed-frequency modeling.
Industrial production is downloaded from the Deutsche Bundesbank with the
[bbk](https://m-muecke.github.io/bbk/) package.

```{r}
#| label: setup
#| message: false
library(ifo)
library(bbk)
library(data.table)
library(ggplot2)
library(lmtest)

theme_ifo <- function(...) {
  theme_minimal(...) +
    theme(
      legend.title = element_blank(),
      legend.position = "top",
      plot.title = element_text(face = "bold"),
      plot.caption = element_text(hjust = 0, vjust = 0, size = 8, margin = margin(10, 0, 0, 0)),
      panel.grid.major.y = element_line(color = "black", linewidth = 0.2),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text = element_text(color = "black"),
      axis.title = element_blank(),
      plot.margin = margin(10, 10, 10, 10)
    )
}
```

### Assembling the data

The ifo Business Climate for Germany comes in long format; we keep the headline *index* values
and reshape the three indicators (climate, situation, expectation) into columns.

```{r}
#| label: ifo-data
survey_long <- setDT(ifo_business("germany"))
survey <- dcast(survey_long[series == "index"], yearmonth ~ indicator, value.var = "value")
head(survey)
```

German industrial production (producing sector excluding construction, calendar and
seasonally adjusted, 2021 = 100) is a single Bundesbank series.
We convert the level index into a **year-on-year growth rate**, the transformation used by
Schröder and Hüfner as their measure of economic activity.

```{r}
#| label: ip-data
ip <- bbk_data("BBDE1", key = "M.DE.Y.BAA1.A2P100000.G.C.I21.A")
ip <- ip[order(date), .(yearmonth = date, ip_growth = 100 * (value / shift(value, 12L) - 1))]

activity <- survey[ip, on = "yearmonth", nomatch = 0L][!is.na(ip_growth)]
```

The production series reaches back to 1991, but the ifo index for Germany as a whole starts in
January 2005, so the overlapping sample runs from
`r format(min(activity$yearmonth), "%B %Y")` to `r format(max(activity$yearmonth), "%B %Y")`
(`r nrow(activity)` months).

### Criterion 1: co-movement

Plotting the ifo climate index against industrial-production growth shows how closely the two
series track each other (both standardized to make them comparable on one axis).

```{r}
#| label: comovement
z <- function(x) (x - mean(x)) / sd(x)
activity[, .(yearmonth, ifo = z(climate), ip = z(ip_growth))] |>
  melt(id.vars = "yearmonth") |>
  ggplot(aes(x = yearmonth, y = value, color = variable)) +
  geom_line() +
  labs(
    title = "ifo Business Climate and industrial production",
    subtitle = "Standardized (z-scores), monthly",
    caption = sprintf(
      "Sources: ifo Business Survey and Deutsche Bundesbank, %s.",
      format(max(activity$yearmonth), "%B %Y")
    )
  ) +
  scale_color_manual(
    values = c(ifo = "darkred", ip = "darkblue"),
    labels = c(ifo = "ifo Business Climate", ip = "Industrial production (YoY growth)")
  ) +
  theme_ifo()
```

The two follow the same broad swings, though they drift apart at times, and the
contemporaneous correlation is moderate rather than tight:

```{r}
#| label: correlation
cor(activity$climate, activity$ip_growth)
```

To see the lead/lag structure we look at the cross-correlation function.
Negative lags describe the ifo index *leading* production growth.
Here it pays to separate the headline climate index into its two ingredients: the assessment
of the **current situation** and the forward-looking **expectations**.
Schröder and Hüfner stress the expectations component, since a genuinely leading signal should
come from what firms anticipate rather than from what they already observe.

```{r}
#| label: crosscorrelation
xcorr <- function(x) {
  cc <- ccf(activity[[x]], activity$ip_growth, lag.max = 12L, plot = FALSE)
  data.table(component = x, lag = as.integer(cc$lag), correlation = as.numeric(cc$acf))
}
xc <- rbindlist(lapply(c("situation", "expectation", "climate"), xcorr))

ggplot(xc, aes(x = lag, y = correlation, color = component)) +
  geom_line() +
  geom_vline(xintercept = 0, linewidth = 0.3) +
  labs(
    title = "Cross-correlation with industrial-production growth",
    subtitle = "Negative lags: the ifo component leads production growth",
    x = "Lag (months)",
    y = "Correlation"
  ) +
  scale_color_manual(
    values = c(climate = "darkred", expectation = "darkblue", situation = "darkgrey"),
    labels = c(
      climate = "Business Climate",
      expectation = "Business Expectations",
      situation = "Business Situation"
    )
  ) +
  theme_ifo() +
  theme(
    axis.title = element_text(),
    panel.grid.major.x = element_line(color = "grey85", linewidth = 0.2)
  )
```

The lag at which each curve peaks orders the three components as the theory suggests:

```{r}
#| label: peak-lags
xc[, .SD[which.max(correlation)], by = component]
```

Expectations peak at or before lag zero, the current-situation assessment clearly after it, and
the headline climate index in between.
The leads are short — a month or two, not a quarter — so co-movement on its own is close to
coincident, but the ordering already hints that the forward-looking component is the one doing
the leading.
To confirm the direction we need a formal test.

### Criterion 2: Granger causality

A leading indicator should carry information about the *future* of the reference series.
The [Granger causality](https://en.wikipedia.org/wiki/Granger_causality) test asks whether past
values of the ifo index improve a forecast of production growth beyond production growth's own
past — and, crucially, whether the reverse also holds.

```{r}
#| label: granger
pval <- function(x) format(signif(x[["Pr(>F)"]][2L], 2L), scientific = FALSE)

forward <- grangertest(ip_growth ~ climate, order = 3L, data = activity)
reverse <- grangertest(climate ~ ip_growth, order = 3L, data = activity)
forward
reverse
```

The result is one-directional: the ifo climate index Granger-causes industrial-production
growth (p = `r pval(forward)`), while production growth does *not* Granger-cause the ifo index
(p = `r pval(reverse)`).
The survey leads the hard data, not the other way around — what a leading indicator should do.

Consistent with the cross-correlation, the forward-looking expectations component gives a
slightly stronger signal than the headline climate index.

```{r}
#| label: granger-expectation
grangertest(ip_growth ~ expectation, order = 3L, data = activity)
```

### Criterion 3: forecast value

Finally, does the ifo index add anything beyond what production growth already says about
itself?
We compare an autoregressive benchmark that predicts production growth from its own three lags
against the same model augmented with three lags of the ifo climate index (the specification
underlying the Granger test above).
The comparison is in-sample: it measures fit, not out-of-sample forecast accuracy.

```{r}
#| label: forecast
dt <- activity[, .(
  ip_growth,
  g1 = shift(ip_growth, 1L),
  g2 = shift(ip_growth, 2L),
  g3 = shift(ip_growth, 3L),
  c1 = shift(climate, 1L),
  c2 = shift(climate, 2L),
  c3 = shift(climate, 3L)
)]
dt <- na.omit(dt)

fit_ar <- lm(ip_growth ~ g1 + g2 + g3, data = dt)
fit_ar_ifo <- lm(ip_growth ~ g1 + g2 + g3 + c1 + c2 + c3, data = dt)

rmse <- function(x) sqrt(mean(residuals(x)^2))
data.table(
  model = c("AR(3)", "AR(3) + ifo"),
  adj_r2 = c(summary(fit_ar)$adj.r.squared, summary(fit_ar_ifo)$adj.r.squared),
  rmse_in_sample = c(rmse(fit_ar), rmse(fit_ar_ifo))
)
```

Adding the ifo index raises the adjusted R² and lowers the in-sample residual error, and the
improvement is statistically significant — the same F test the Granger procedure ran above,
here in its regression form:

```{r}
#| label: forecast-test
anova(fit_ar, fit_ar_ifo)
```

### An out-of-sample check

In-sample fit does not guarantee a better forecast. Lehmann's review also suggests that
manufacturing expectations should be a closer match for industrial production than an
economy-wide indicator. We test that directly with an expanding window: after an initial ten
years of data, each model is re-estimated monthly before predicting the next observation.

```{r}
#| label: rolling-forecast
germany_expectation <- survey_long[
  series == "balance" & indicator == "expectation",
  .(yearmonth, de_expectation = value)
]
manufacturing_expectation <- setDT(ifo_business("sectors"))[
  sector == "manufacturing" & series == "balance" & indicator == "expectation",
  .(yearmonth, mfg_expectation = value)
]
forecast_data <- mergelist(
  list(
    activity[, .(yearmonth, ip_growth)],
    germany_expectation,
    manufacturing_expectation
  ),
  on = "yearmonth",
  how = "inner"
)

lag_variables <- setdiff(names(forecast_data), "yearmonth")
lags <- 1:3
lag_columns <- paste0(rep(lag_variables, each = length(lags)), lags)
forecast_data[, (lag_columns) := shift(.SD, lags), .SDcols = lag_variables]

models <- list(
  "AR(3)" = paste0("ip_growth", 1:3),
  "Germany expectations" = c(paste0("ip_growth", 1:3), paste0("de_expectation", 1:3)),
  "Manufacturing expectations" = c(paste0("ip_growth", 1:3), paste0("mfg_expectation", 1:3))
)

required <- c("ip_growth", unique(unlist(models)))
forecast_data <- na.omit(forecast_data, cols = required)
initial_window <- 120L
forecast_months <- initial_window + seq_len(nrow(forecast_data) - initial_window)

predictions <- rbindlist(lapply(names(models), function(model) {
  formula <- reformulate(models[[model]], response = "ip_growth")
  predicted <- vapply(
    forecast_months,
    function(month) {
      fit <- lm(formula, data = forecast_data[seq_len(month - 1L)])
      predict(fit, newdata = forecast_data[month])
    },
    numeric(1L)
  )
  data.table(
    model,
    yearmonth = forecast_data$yearmonth[forecast_months],
    actual = forecast_data$ip_growth[forecast_months],
    predicted
  )
}))

forecast_scores <- predictions[,
  .(months = .N, rmse = sqrt(mean((actual - predicted)^2))),
  by = model
]
forecast_scores[,
  improvement_over_ar := 100 * (rmse[model == "AR(3)"] - rmse) / rmse[model == "AR(3)"]
]
forecast_scores[, let(
  rmse = round(rmse, 2L),
  improvement_over_ar = round(improvement_over_ar, 1L)
)]
forecast_scores[]
```

The Germany-wide expectations balance barely improves on the benchmark. Manufacturing
expectations reduce the forecast error more clearly, consistent with Lehmann's finding that
sector-specific indicators are better suited to sector-specific targets. This remains a
pseudo-out-of-sample exercise because it uses the latest data vintage.

### Caveats

**The pandemic dominates the sample.** The co-movement chart shows production growth swinging by
more than five standard deviations in 2020 and 2021, and those two dozen observations carry a lot
of weight in every statistic on this page. Before drawing strong conclusions it is worth re-running
the exercise on a pre-2020 sample, or adding pandemic dummies, to check that the ordering
survives.

**These results use the *final vintage* of industrial production.** In real time the ifo index
is even more valuable, because the official production figures for a given month are published
with a delay of several weeks and are subsequently revised — precisely the gap a timely survey
fills. A real-time evaluation may therefore differ from the pseudo-out-of-sample results above;
see
[Lehmann and Reif (2021)](https://link.springer.com/article/10.1007/s41549-021-00055-5) for a
rigorous real-time horse-race. The exercise above is meant to illustrate the mechanics of
assessing a leading indicator, not to settle its forecasting record.
