---
title: "Evaluating the ifo Business Climate as a leading indicator"
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  dpi = 300L
)
```

Firms report their current situation and expectations before official statistics become
available, so the ifo Business Climate may reveal changes in economic activity earlier.
We test this claim by adapting the evaluation framework of
[Hüfner and Schröder (2002)](https://ftp.zew.de/pub/zew-docs/dp/dp0256.pdf).
This article revisits the main qualitative tests with current data rather than reproducing the
original samples or forecast tables.
[Lehmann (2023)](https://link.springer.com/article/10.1007/s41549-022-00079-5) provides a
broader review of the forecasting evidence from the ifo Business Survey.

Following [Fritsche and Stephan (2000)](https://www.diw.de/documents/publikationen/73/diw_01.c.38582.de/dp207.pdf),
we ask whether the ifo index satisfies three criteria of a good leading indicator:

1. **Co-movement.** The indicator should move with industrial production.
2. **Predictive content.** The relationship should be statistically significant and stable
   over time, and the indicator should improve on a simple autoregressive model of activity.
3. **Forecast value.** Including the indicator should improve out-of-sample forecasts relative
   to a naive benchmark.

We use the ifo Business Climate together with the German industrial production index.
Both series are monthly, so no mixed-frequency modeling is needed.
We download industrial production from the Deutsche Bundesbank with the
[bbk](https://m-muecke.github.io/bbk/) package.

```{r}
#| label: setup
#| message: false
library(ifo)
library(bbk)
library(data.table)
library(ggplot2)

theme_ifo <- function(...) {
  theme_minimal(...) +
    theme(
      legend.title = element_blank(),
      legend.position = "top",
      plot.title = element_text(face = "bold"),
      plot.caption = element_text(hjust = 0, vjust = 0, size = 8, margin = margin(10, 0, 0, 0)),
      panel.grid.major.y = element_line(color = "black", linewidth = 0.2),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text = element_text(color = "black"),
      axis.title = element_blank(),
      plot.margin = margin(10, 10, 10, 10)
    )
}
```

### Assembling the data

The ifo Business Climate for Germany comes in long format. We keep the headline *index* values
and reshape climate, situation, and expectations into columns.

```{r}
#| label: ifo-data
survey_long <- setDT(ifo_business("germany"))
survey <- dcast(survey_long[series == "index"], yearmonth ~ indicator, value.var = "value")
head(survey)
```

German industrial production (producing sector excluding construction, calendar and
seasonally adjusted, 2021 = 100) is a single Bundesbank series.
We convert the level index into a **year-on-year growth rate**, the transformation used by
Hüfner and Schröder as their measure of economic activity.

```{r}
#| label: ip-data
ip <- bbk_data("BBDE1", key = "M.DE.Y.BAA1.A2P100000.G.C.I21.A") |>
  setorder(date) |>
  _[, .(yearmonth = date, ip_growth = 100 * (value / shift(value, 12L) - 1))]

activity <- survey[ip, on = "yearmonth", nomatch = 0L]
```

The production series reaches back to 1991, but the ifo index for Germany as a whole starts in
January 2005, so the overlapping sample runs from
`r format(min(activity$yearmonth), "%B %Y")` to `r format(max(activity$yearmonth), "%B %Y")`
(`r nrow(activity)` months).

### Criterion 1: co-movement

Plotting the ifo climate index against industrial-production growth shows how closely the two
series track each other (both standardized to make them comparable on one axis).

```{r}
#| label: comovement
z <- function(x) (x - mean(x)) / sd(x)
activity[, .(yearmonth, ifo = z(climate), ip = z(ip_growth))] |>
  melt(id.vars = "yearmonth") |>
  ggplot(aes(x = yearmonth, y = value, color = variable)) +
  geom_line() +
  labs(
    title = "ifo Business Climate and industrial production",
    subtitle = "Standardized (z-scores), monthly",
    caption = sprintf(
      "Sources: ifo Business Survey and Deutsche Bundesbank, %s.",
      format(max(activity$yearmonth), "%B %Y")
    )
  ) +
  scale_color_manual(
    values = c(ifo = "darkred", ip = "darkblue"),
    labels = c(ifo = "ifo Business Climate", ip = "Industrial production (YoY growth)")
  ) +
  theme_ifo()
```

The two follow the same broad swings, though they drift apart at times, and the
contemporaneous correlation is moderate rather than tight:

```{r}
#| label: correlation
cor(activity$climate, activity$ip_growth)
```

To examine the lead-lag structure, we use the cross-correlation function.
Negative lags describe the ifo index *leading* production growth.
We separate the headline climate index into the assessment of the **current situation** and the
forward-looking **expectations**.
Hüfner and Schröder stress the expectations component, since a leading signal should
come from what firms anticipate rather than from what they already observe.

```{r}
#| label: crosscorrelation
cross_correlation <- function(dt) {
  dt |>
    melt(
      id.vars = "ip_growth",
      measure.vars = c("situation", "expectation", "climate"),
      variable.name = "component",
      value.name = "ifo_value",
      variable.factor = FALSE
    ) |>
    _[,
      {
        cc <- ccf(ifo_value, ip_growth, lag.max = 12L, plot = FALSE)
        .(lag = as.integer(cc$lag), correlation = as.numeric(cc$acf))
      },
      by = component
    ]
}

plot_cross_correlation <- function(xc, subtitle) {
  ggplot(xc, aes(x = lag, y = correlation, color = component)) +
    geom_line() +
    geom_vline(xintercept = 0, linewidth = 0.3) +
    labs(
      title = "Cross-correlation with industrial-production growth",
      subtitle = subtitle,
      x = "Lag (months)",
      y = "Correlation",
      caption = sprintf(
        "Sources: ifo Business Survey and Deutsche Bundesbank, %s.",
        format(max(activity$yearmonth), "%B %Y")
      )
    ) +
    scale_color_manual(
      values = c(climate = "darkred", expectation = "darkblue", situation = "darkgrey"),
      labels = c(
        climate = "Business Climate",
        expectation = "Business Expectations",
        situation = "Business Situation"
      )
    ) +
    theme_ifo() +
    theme(
      axis.title = element_text(),
      panel.grid.major.x = element_line(color = "grey85", linewidth = 0.2)
    )
}

xc <- cross_correlation(activity)
plot_cross_correlation(xc, "Negative lags: the ifo component leads production growth")
```

The lag at which each curve peaks orders the three components as the theory suggests:

```{r}
#| label: peak-lags
xc[, .SD[which.max(correlation)], by = component]
```

The peak correlation occurs at lag
`r xc[component == "expectation", lag[which.max(correlation)]]` for expectations,
`r xc[component == "climate", lag[which.max(correlation)]]` for the climate index, and
`r xc[component == "situation", lag[which.max(correlation)]]` for the situation assessment.
Because negative lags indicate a lead over production, only expectations peak ahead of
production, and the lead is short: the correlation is nearly flat between lag -1 and lag 0.

### Matching the transformation

The short lead partly reflects how the two series are transformed.
[Lehmann (2023)](https://link.springer.com/article/10.1007/s41549-022-00079-5) argues,
following Abberger and Nierhaus (2007, *ifo Schnelldienst* 60(5), 25-30), that survey balances
are stationary by construction and are naturally compared in levels with the cyclical component
of activity. A growth rate is an asymmetric filter that shifts the turning points of production
backwards in time, so comparing survey levels with production growth disguises the lead. The
remedy is to apply the same transformation to the survey series, here the change over the same
month of the previous year.

```{r}
#| label: crosscorrelation-diff
diff12 <- function(x) x - shift(x, 12L)
activity_diff <- activity[, .(
  ip_growth,
  situation = diff12(situation),
  expectation = diff12(expectation),
  climate = diff12(climate)
)] |>
  na.omit()

xc_diff <- cross_correlation(activity_diff)
plot_cross_correlation(xc_diff, "ifo components as twelve-month differences")
```

```{r}
#| label: peak-lags-diff
xc_diff[, .SD[which.max(correlation)], by = component]
plateau <- xc_diff[component == "expectation"][correlation >= 0.95 * max(correlation), range(lag)]
```

With matched transformations, expectations peak at lag
`r xc_diff[component == "expectation", lag[which.max(correlation)]]`, and the correlation
stays within five percent of that peak from lag `r plateau[1]` to lag `r plateau[2]`.
The lead is a plateau of several months rather than the single month found in levels.
The situation assessment and the climate index peak at lag
`r xc_diff[component == "situation", lag[which.max(correlation)]]` and
`r xc_diff[component == "climate", lag[which.max(correlation)]]` with correlations of
`r sprintf("%.2f", xc_diff[component == "situation", max(correlation)])` and
`r sprintf("%.2f", xc_diff[component == "climate", max(correlation)])`, so they become
coincident indicators that track production growth more closely than in levels.
The remaining sections keep the survey series in levels, as Hüfner and Schröder do. The
regressions there include several lags of each series, which absorbs part of the phase shift.
A formal test can determine whether the lead holds after accounting for the history of
production itself.

### Criterion 2: Granger causality

A leading indicator should carry information about the *future* of the reference series.
The [Granger causality](https://en.wikipedia.org/wiki/Granger_causality) test asks whether past
values of the ifo index improve a forecast of production growth beyond production growth's own
past. We also test the reverse direction: an indicator that merely reacts to production would
be predicted by it. This tests predictive information, not an economic causal mechanism.
We use three lags of each variable. The conclusion does not hinge on this choice: the Schwarz
criterion prefers two lags and the Akaike criterion considerably more, and the test results
are the same for any order between them.

```{r}
#| label: granger
pval <- function(x) format(signif(x[["Pr(>F)"]][2L], 2L), scientific = FALSE)

forward <- lmtest::grangertest(ip_growth ~ climate, order = 3L, data = activity)
reverse <- lmtest::grangertest(climate ~ ip_growth, order = 3L, data = activity)
forward
reverse
```

The predictive relationship is one-directional. Lags of the ifo climate improve the production
model (p = `r pval(forward)`), while lags of production do not improve the climate model
(p = `r pval(reverse)`).

Consistent with the cross-correlation, the forward-looking expectations component gives a
slightly stronger signal than the headline climate index.

```{r}
#| label: granger-expectation
lmtest::grangertest(ip_growth ~ expectation, order = 3L, data = activity)
```

### Criterion 2: in-sample fit

The second criterion also asks the indicator to improve on a simple autoregressive model of
activity. Does the ifo index add anything beyond what production growth already says about
itself? We compare an autoregressive benchmark that predicts production growth from its own
three lags against the same model augmented with three lags of the ifo climate index (the
specification underlying the Granger test above).
The comparison is in-sample: it measures fit, not out-of-sample forecast accuracy.

```{r}
#| label: forecast
dt <- activity |>
  _[, .(
    ip_growth,
    g1 = shift(ip_growth, 1L),
    g2 = shift(ip_growth, 2L),
    g3 = shift(ip_growth, 3L),
    c1 = shift(climate, 1L),
    c2 = shift(climate, 2L),
    c3 = shift(climate, 3L)
  )] |>
  na.omit()

fit_ar <- lm(ip_growth ~ g1 + g2 + g3, data = dt)
fit_ar_ifo <- lm(ip_growth ~ g1 + g2 + g3 + c1 + c2 + c3, data = dt)

rmse <- function(x) sqrt(mean(residuals(x)^2))
data.table(
  model = c("AR(3)", "AR(3) + ifo"),
  adj_r2 = c(summary(fit_ar)$adj.r.squared, summary(fit_ar_ifo)$adj.r.squared),
  rmse_in_sample = c(rmse(fit_ar), rmse(fit_ar_ifo))
)
```

Adding the climate index raises adjusted R² from
`r sprintf("%.3f", summary(fit_ar)$adj.r.squared)` to
`r sprintf("%.3f", summary(fit_ar_ifo)$adj.r.squared)` and lowers in-sample RMSE from
`r sprintf("%.2f", rmse(fit_ar))` to `r sprintf("%.2f", rmse(fit_ar_ifo))`.
The nested-model F test rejects the three climate coefficients being jointly zero
(p = `r pval(forward)`). This is the same test used in the forward Granger comparison above:

```{r}
#| label: forecast-test
anova(fit_ar, fit_ar_ifo)
```

### Criterion 3: forecast value

In-sample fit does not guarantee a better forecast, so the third criterion asks whether the
indicator improves out-of-sample forecasts. We switch from the index values used above to the
balances: the vintage workbooks provide balances only, and Hüfner and Schröder also work with
the expectations balance. We compare the Germany-wide climate, situation, and expectations
balances in an expanding-window design. We also include manufacturing expectations to test
whether a more closely matched survey series improves the forecast, as suggested by Lehmann's
review. For manufacturing expectations, we use the value first published for each month. This
prevents later survey revisions from entering earlier forecasts. Starting with ten years of
data, we re-estimate each model monthly and predict the next observation.

```{r}
#| label: rolling-forecast
germany_components <- survey_long |>
  _[series == "balance"] |>
  dcast(yearmonth ~ indicator, value.var = "value") |>
  setnames(
    c("climate", "situation", "expectation"),
    c("de_climate", "de_situation", "de_expectation")
  )
manufacturing_expectation <- setDT(ifo_vintage("manufacturing"))[
  vintage == yearmonth & series == "balance" & indicator == "expectation",
  .(yearmonth, mfg_expectation = value)
]
forecast_data <- mergelist(
  list(
    activity[, .(yearmonth, ip_growth)],
    germany_components,
    manufacturing_expectation
  ),
  on = "yearmonth",
  how = "inner"
)

lag_variables <- setdiff(names(forecast_data), "yearmonth")
lags <- 1:3
lag_columns <- paste0(rep(lag_variables, each = length(lags)), lags)
forecast_data[, (lag_columns) := shift(.SD, lags), .SDcols = lag_variables]

models <- list(
  "AR(3)" = paste0("ip_growth", lags),
  "Germany climate" = c(paste0("ip_growth", lags), paste0("de_climate", lags)),
  "Germany situation" = c(paste0("ip_growth", lags), paste0("de_situation", lags)),
  "Germany expectations" = c(paste0("ip_growth", lags), paste0("de_expectation", lags)),
  "Manufacturing expectations" = c(paste0("ip_growth", lags), paste0("mfg_expectation", lags))
)

required <- c("ip_growth", unique(unlist(models)))
forecast_data <- na.omit(forecast_data, cols = required)
initial_window <- 120L
forecast_months <- initial_window + seq_len(nrow(forecast_data) - initial_window)

roll_forecast <- function(predictors, model) {
  formula <- reformulate(predictors, response = "ip_growth")
  predicted <- vapply(
    forecast_months,
    function(month) {
      fit <- lm(formula, data = forecast_data[seq_len(month - 1L)])
      predict(fit, newdata = forecast_data[month])
    },
    numeric(1L)
  )
  data.table(
    model,
    yearmonth = forecast_data$yearmonth[forecast_months],
    actual = forecast_data$ip_growth[forecast_months],
    predicted
  )
}

predictions <- rbindlist(Map(roll_forecast, predictors = models, model = names(models)))

forecast_scores <- predictions |>
  _[, .(months = .N, rmse = sqrt(mean((actual - predicted)^2))), by = model] |>
  _[, improvement_over_ar := 100 * (rmse[model == "AR(3)"] - rmse) / rmse[model == "AR(3)"]] |>
  _[, let(rmse = round(rmse, 2L), improvement_over_ar = round(improvement_over_ar, 1L))]
forecast_scores[]
```

Across `r forecast_scores[model == "AR(3)", months]` forecasts, Germany-wide expectations lower
RMSE by `r forecast_scores[model == "Germany expectations", improvement_over_ar]`% relative to
the AR benchmark, slightly more than the climate balance at
`r forecast_scores[model == "Germany climate", improvement_over_ar]`%. The situation balance
raises RMSE by `r abs(forecast_scores[model == "Germany situation", improvement_over_ar])`%.
Manufacturing expectations lower it by
`r forecast_scores[model == "Manufacturing expectations", improvement_over_ar]`%.
The forecast ranking therefore agrees with the lead-lag analysis, although the differences
between the Germany-wide components are small. The closer sectoral match appears to matter, but
industrial production is broader than manufacturing alone. This remains a pseudo-out-of-sample
exercise because industrial production and the Germany-wide balances use their latest vintages.
The manufacturing series uses first-release values.

Are these gains statistically distinguishable from the benchmark? A
[Diebold and Mariano (1995)](https://doi.org/10.1080/07350015.1995.10524599) test on the
squared-error differentials would be conservative here because the models are nested: each
survey model contains the AR(3) benchmark, so its extra coefficients add estimation noise to
its forecasts. [Clark and West (2007)](https://doi.org/10.1016/j.jeconom.2006.05.023) correct
for this by adding the squared difference between the two forecasts back to the loss
differential. The adjusted statistic is approximately standard normal, and the test is
one-sided: does the survey model improve on the benchmark?

```{r}
#| label: cw-test
cw_scores <- predictions |>
  _[, .(yearmonth, model, actual, predicted)] |>
  _[, benchmark := predicted[model == "AR(3)"], by = yearmonth] |>
  _[
    model != "AR(3)",
    {
      f <- (actual - benchmark)^2 - (actual - predicted)^2 + (benchmark - predicted)^2
      cw <- mean(f) / sqrt(var(f) / .N)
      .(cw_statistic = cw, p_value = pnorm(-cw))
    },
    by = model
  ]
cw_scores[, .(
  model,
  cw_statistic = round(cw_statistic, 2L),
  p_value = round(p_value, 2L)
)]
```

Only manufacturing expectations clear the 5% level
(p = `r cw_scores[model == "Manufacturing expectations", sprintf("%.2f", p_value)]`).
Germany-wide expectations are borderline
(p = `r cw_scores[model == "Germany expectations", sprintf("%.2f", p_value)]`), and the climate and
situation balances are clearly insignificant. The evidence for added predictive content is therefore
concentrated in the survey series that matches the reference variable most closely. Whether that
evidence rests on a few extreme months is the next question.

### When do the survey indicators help?

The second criterion also requires the relationship to be stable over time, and the average
RMSE hides when the forecast gains occur. We accumulate the monthly difference in
squared errors between the AR benchmark and each model with expectations. The line rises when
the survey model produces the smaller error and falls when the benchmark does.

```{r}
#| label: forecast-stability
forecast_gain <- predictions |>
  _[, .(yearmonth, model, error = (actual - predicted)^2)] |>
  _[, benchmark_error := error[model == "AR(3)"], by = yearmonth] |>
  _[model %in% c("Germany expectations", "Manufacturing expectations")] |>
  setorder(model, yearmonth) |>
  _[, .(yearmonth, cumulative_gain = cumsum(benchmark_error - error)), by = model]

ggplot(forecast_gain, aes(x = yearmonth, y = cumulative_gain, color = model)) +
  annotate(
    "rect",
    xmin = as.Date("2020-01-01"),
    xmax = as.Date("2021-12-31"),
    ymin = -Inf,
    ymax = Inf,
    fill = "grey90",
    color = NA
  ) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_line() +
  labs(
    title = "Forecast gains are concentrated in the pandemic",
    subtitle = "Cumulative reduction in squared error relative to AR(3)",
    y = "Cumulative squared-error reduction",
    caption = sprintf(
      "Shaded area: January 2020 to December 2021.\nSources: ifo Business Survey and Deutsche Bundesbank, %s.",
      format(max(activity$yearmonth), "%B %Y")
    )
  ) +
  scale_color_manual(
    values = c("Germany expectations" = "darkred", "Manufacturing expectations" = "darkblue")
  ) +
  theme_ifo() +
  theme(axis.title.y = element_text())
```

### Caveats

**The pandemic has an outsized influence on the sample.** Most of the cumulative forecast gain
appears in 2020 and 2021. After 2022, Germany-wide expectations give back most of that
advantage and manufacturing expectations about half of it. The full-sample RMSE therefore does
not show a stable forecasting advantage. A pre-2020 estimation or pandemic
indicators would provide a stricter sensitivity check.

**Industrial production and the Germany-wide ifo series use their latest vintages.** Manufacturing
expectations use first-release values. The ifo survey appears during its reference month, while
official production figures appear several weeks later and are revised. A fully real-time nowcast
would need release-date-aligned production vintages and could not use unpublished production lags.
See
[Lehmann and Reif (2021)](https://link.springer.com/article/10.1007/s41549-021-00055-5) for a
real-time comparison. This exercise demonstrates how to assess a leading indicator. It does not
settle the ifo survey's forecasting record.
