---
title: "The ifo index as a leading indicator"
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  dpi = 300L
)
```

The ifo Business Climate Index is widely used as a *leading indicator* for the German
economy: because firms are surveyed about their current situation and their expectations
before the official statistics are compiled, the index is meant to signal turning points in
economic activity ahead of the hard data.
This article puts that claim to the test by (loosely) replicating
[Schröder and Hüfner (2002)](https://ftp.zew.de/pub/zew-docs/dp/dp0256.pdf), who study how
useful sentiment indicators are for forecasting German economic activity.

Following [Fritsche and Stephan (2002)](https://www.diw.de/documents/publikationen/73/diw_01.c.38193.de/dp241.pdf),
we ask whether the ifo index satisfies three criteria of a good leading indicator:

1. **Co-movement** with the reference series (industrial production).
2. A statistically significant, one-directional relationship — the indicator should help
   predict activity, not merely react to it.
3. **Forecast value** — it should improve a forecast beyond what the activity series predicts
   about itself.

We use the ifo Business Climate together with the German industrial production index, which is
monthly (like the ifo survey) and therefore needs no mixed-frequency modelling.
Industrial production is downloaded from the Deutsche Bundesbank with the
[bbk](https://m-muecke.github.io/bbk/) package.

```{r}
#| label: setup
#| message: false
library(ifo)
library(bbk)
library(data.table)
library(ggplot2)

theme_ifo <- function(...) {
  theme_minimal(...) +
    theme(
      legend.title = element_blank(),
      legend.position = "top",
      plot.title = element_text(face = "bold"),
      plot.caption = element_text(hjust = 0, vjust = 0, size = 8, margin = margin(10, 0, 0, 0)),
      panel.grid.major.y = element_line(color = "black", linewidth = 0.2),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text = element_text(color = "black"),
      plot.margin = margin(10, 10, 10, 10)
    )
}
```

### Assembling the data

The ifo Business Climate for Germany comes in long format; we keep the headline *index* values
and reshape the three indicators (climate, situation, expectation) into columns.

```{r}
#| label: ifo-data
ifo <- setDT(ifo_business("germany"))
ifo <- ifo[series == "index"]
ifo <- dcast(ifo, yearmonth ~ indicator, value.var = "value")
head(ifo)
```

German industrial production (producing sector excluding construction, calendar and
seasonally adjusted, 2021 = 100) is a single Bundesbank series.
We convert the level index into a **year-on-year growth rate**, the transformation used by
Schröder and Hüfner as their measure of economic activity.

```{r}
#| label: ip-data
ip <- bbk_data("BBDE1", key = "M.DE.Y.BAA1.A2P100000.G.C.I21.A")
ip <- ip[order(date), .(yearmonth = date, ip = value)]
ip[, ip_growth := 100 * (ip / shift(ip, 12L) - 1)]

activity <- merge(ifo, ip, by = "yearmonth")[!is.na(ip_growth)]
```

### Criterion 1: co-movement

Plotting the ifo climate index against industrial-production growth shows how tightly the two
series move together (both standardised to make them comparable on one axis).

```{r}
#| label: comovement
#| dpi: 300
z <- function(x) (x - mean(x)) / sd(x)
activity[, .(yearmonth, ifo = z(climate), ip = z(ip_growth))] |>
  melt(id.vars = "yearmonth") |>
  ggplot(aes(x = yearmonth, y = value, color = variable)) +
  geom_line() +
  labs(
    title = "ifo Business Climate and industrial production",
    subtitle = "Standardised (z-scores), monthly",
    caption = sprintf(
      "Sources: ifo Business Survey and Deutsche Bundesbank, %s.",
      format(max(activity$yearmonth), "%B %Y")
    )
  ) +
  scale_color_manual(
    values = c(ifo = "darkred", ip = "darkblue"),
    labels = c("ifo Business Climate", "Industrial production (YoY growth)")
  ) +
  theme_ifo()
```

The contemporaneous correlation is high:

```{r}
#| label: correlation
cor(activity$climate, activity$ip_growth)
```

To see the lead/lag structure we look at the cross-correlation function.
Negative lags describe the ifo index *leading* production growth.
Here it pays to separate the headline climate index into its two ingredients: the assessment
of the **current situation** and the forward-looking **expectations**.
Schröder and Hüfner stress the expectations component, since a genuinely leading signal should
come from what firms anticipate rather than from what they already observe.

```{r}
#| label: crosscorrelation
#| dpi: 300
xcorr <- function(x) {
  cc <- ccf(activity[[x]], activity$ip_growth, lag.max = 12L, plot = FALSE)
  data.table(component = x, lag = as.integer(cc$lag), correlation = as.numeric(cc$acf))
}

rbindlist(lapply(c("situation", "expectation", "climate"), xcorr)) |>
  ggplot(aes(x = lag, y = correlation, color = component)) +
  geom_line() +
  geom_vline(xintercept = 0, linewidth = 0.3) +
  labs(
    title = "Cross-correlation with industrial-production growth",
    subtitle = "Negative lags: the ifo component leads production growth"
  ) +
  scale_color_manual(
    values = c(situation = "darkgrey", expectation = "darkblue", climate = "darkred"),
    labels = c("Business Climate", "Business Expectations", "Business Situation")
  ) +
  theme_ifo() +
  theme(panel.grid.major.x = element_line(color = "grey85", linewidth = 0.2))
```

The three curves peak at different lags: expectations peak at a *negative* lag (leading
production growth), the current-situation assessment peaks at a *positive* lag (lagging it), and
the headline climate index sits between the two.
So co-movement on its own is broadly coincident, but it already hints that the forward-looking
component leads.
To confirm the direction we need a formal test.

### Criterion 2: Granger causality

A leading indicator should carry information about the *future* of the reference series.
The [Granger causality](https://en.wikipedia.org/wiki/Granger_causality) test asks whether past
values of the ifo index improve a forecast of production growth beyond production growth's own
past — and, crucially, whether the reverse also holds.

```{r}
#| label: granger
lmtest::grangertest(ip_growth ~ climate, order = 3L, data = activity)
lmtest::grangertest(climate ~ ip_growth, order = 3L, data = activity)
```

The result is clean and one-directional: the ifo climate index Granger-causes
industrial-production growth (p < 0.001), while production growth does *not* Granger-cause the
ifo index.
The survey leads the hard data, not the other way around — exactly what a leading indicator
should do.

Consistent with the cross-correlation, the forward-looking expectations component gives an even
stronger signal than the headline climate index.

```{r}
#| label: granger-expectation
lmtest::grangertest(ip_growth ~ expectation, order = 3L, data = activity)
```

### Criterion 3: forecast value

Finally, does the ifo index actually improve a forecast?
We compare an autoregressive benchmark that predicts production growth from its own three lags
against the same model augmented with three lags of the ifo climate index (the specification
underlying the Granger test above).

```{r}
#| label: forecast
dt <- activity[, .(
  ip_growth,
  g1 = shift(ip_growth, 1L), g2 = shift(ip_growth, 2L), g3 = shift(ip_growth, 3L),
  c1 = shift(climate, 1L), c2 = shift(climate, 2L), c3 = shift(climate, 3L)
)]
dt <- dt[complete.cases(dt)]

ar <- lm(ip_growth ~ g1 + g2 + g3, data = dt)
ar_ifo <- lm(ip_growth ~ g1 + g2 + g3 + c1 + c2 + c3, data = dt)

rmse <- function(m) sqrt(mean(residuals(m)^2))
data.table(
  model = c("AR(3)", "AR(3) + ifo"),
  adj_r2 = c(summary(ar)$adj.r.squared, summary(ar_ifo)$adj.r.squared),
  rmse = c(rmse(ar), rmse(ar_ifo))
)
```

Adding the ifo index lowers the in-sample forecast error and raises the adjusted R², and the
improvement is statistically significant:

```{r}
#| label: forecast-test
anova(ar, ar_ifo)
```

### Caveat: real-time versus final data

These results use the *final vintage* of industrial production. In real time the ifo index is
even more valuable, because the official production figures for a given month are published with
a delay of several weeks and are subsequently revised — precisely the gap a timely survey fills.
Conversely, an out-of-sample, real-time evaluation would show smaller gains than the in-sample
fit here; see
[Lehmann and Reif (2021)](https://link.springer.com/article/10.1007/s41549-021-00055-5) for a
rigorous real-time horse-race. The exercise above is meant to illustrate the mechanics of
assessing a leading indicator, not to settle its forecasting record.
