---
title: "The ifo Business Climate as a leading indicator"
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  dpi = 300L
)
```

Firms report their current situation and expectations before official statistics become
available, so the ifo Business Climate may reveal changes in economic activity earlier.
We test this claim by adapting the evaluation framework of
[Schröder and Hüfner (2002)](https://ftp.zew.de/pub/zew-docs/dp/dp0256.pdf).
This article revisits the main qualitative tests with current data rather than reproducing the
original samples or forecast tables.
[Lehmann (2023)](https://link.springer.com/article/10.1007/s41549-022-00079-5) provides a
broader review of the forecasting evidence from the ifo Business Survey.

Following [Fritsche and Stephan (2000)](https://www.diw.de/documents/publikationen/73/diw_01.c.38582.de/dp207.pdf),
we ask whether the ifo index satisfies three criteria of a good leading indicator:

1. **Co-movement.** The indicator should move with industrial production.
2. **Predictive direction.** Past values of the indicator should help predict activity, while
   past values of activity should not predict the indicator.
3. **Forecast value.** The indicator should improve a forecast beyond what the activity series
   predicts about itself.

We use the ifo Business Climate together with the German industrial production index.
Both series are monthly, so no mixed-frequency modeling is needed.
We download industrial production from the Deutsche Bundesbank with the
[bbk](https://m-muecke.github.io/bbk/) package.

```{r}
#| label: setup
#| message: false
library(ifo)
library(bbk)
library(data.table)
library(ggplot2)

theme_ifo <- function(...) {
  theme_minimal(...) +
    theme(
      legend.title = element_blank(),
      legend.position = "top",
      plot.title = element_text(face = "bold"),
      plot.caption = element_text(hjust = 0, vjust = 0, size = 8, margin = margin(10, 0, 0, 0)),
      panel.grid.major.y = element_line(color = "black", linewidth = 0.2),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text = element_text(color = "black"),
      axis.title = element_blank(),
      plot.margin = margin(10, 10, 10, 10)
    )
}
```

### Assembling the data

The ifo Business Climate for Germany comes in long format. We keep the headline *index* values
and reshape climate, situation, and expectations into columns.

```{r}
#| label: ifo-data
survey_long <- setDT(ifo_business("germany"))
survey <- dcast(survey_long[series == "index"], yearmonth ~ indicator, value.var = "value")
head(survey)
```

German industrial production (producing sector excluding construction, calendar and
seasonally adjusted, 2021 = 100) is a single Bundesbank series.
We convert the level index into a **year-on-year growth rate**, the transformation used by
Schröder and Hüfner as their measure of economic activity.

```{r}
#| label: ip-data
ip <- bbk_data("BBDE1", key = "M.DE.Y.BAA1.A2P100000.G.C.I21.A") |>
  setorder(date) |>
  _[, .(yearmonth = date, ip_growth = 100 * (value / shift(value, 12L) - 1))]

activity <- survey[ip, on = "yearmonth", nomatch = 0L]
```

The production series reaches back to 1991, but the ifo index for Germany as a whole starts in
January 2005, so the overlapping sample runs from
`r format(min(activity$yearmonth), "%B %Y")` to `r format(max(activity$yearmonth), "%B %Y")`
(`r nrow(activity)` months).

### Criterion 1: co-movement

Plotting the ifo climate index against industrial-production growth shows how closely the two
series track each other (both standardized to make them comparable on one axis).

```{r}
#| label: comovement
z <- function(x) (x - mean(x)) / sd(x)
activity[, .(yearmonth, ifo = z(climate), ip = z(ip_growth))] |>
  melt(id.vars = "yearmonth") |>
  ggplot(aes(x = yearmonth, y = value, color = variable)) +
  geom_line() +
  labs(
    title = "ifo Business Climate and industrial production",
    subtitle = "Standardized (z-scores), monthly",
    caption = sprintf(
      "Sources: ifo Business Survey and Deutsche Bundesbank, %s.",
      format(max(activity$yearmonth), "%B %Y")
    )
  ) +
  scale_color_manual(
    values = c(ifo = "darkred", ip = "darkblue"),
    labels = c(ifo = "ifo Business Climate", ip = "Industrial production (YoY growth)")
  ) +
  theme_ifo()
```

The two follow the same broad swings, though they drift apart at times, and the
contemporaneous correlation is moderate rather than tight:

```{r}
#| label: correlation
cor(activity$climate, activity$ip_growth)
```

To examine the lead-lag structure, we use the cross-correlation function.
Negative lags describe the ifo index *leading* production growth.
We separate the headline climate index into the assessment of the **current situation** and the
forward-looking **expectations**.
Schröder and Hüfner stress the expectations component, since a leading signal should
come from what firms anticipate rather than from what they already observe.

```{r}
#| label: crosscorrelation
xc <- activity |>
  melt(
    id.vars = "ip_growth",
    measure.vars = c("situation", "expectation", "climate"),
    variable.name = "component",
    value.name = "ifo_value",
    variable.factor = FALSE
  ) |>
  _[,
    {
      cc <- ccf(ifo_value, ip_growth, lag.max = 12L, plot = FALSE)
      .(lag = as.integer(cc$lag), correlation = as.numeric(cc$acf))
    },
    by = component
  ]

ggplot(xc, aes(x = lag, y = correlation, color = component)) +
  geom_line() +
  geom_vline(xintercept = 0, linewidth = 0.3) +
  labs(
    title = "Cross-correlation with industrial-production growth",
    subtitle = "Negative lags: the ifo component leads production growth",
    x = "Lag (months)",
    y = "Correlation",
    caption = sprintf(
      "Sources: ifo Business Survey and Deutsche Bundesbank, %s.",
      format(max(activity$yearmonth), "%B %Y")
    )
  ) +
  scale_color_manual(
    values = c(climate = "darkred", expectation = "darkblue", situation = "darkgrey"),
    labels = c(
      climate = "Business Climate",
      expectation = "Business Expectations",
      situation = "Business Situation"
    )
  ) +
  theme_ifo() +
  theme(
    axis.title = element_text(),
    panel.grid.major.x = element_line(color = "grey85", linewidth = 0.2)
  )
```

The lag at which each curve peaks orders the three components as the theory suggests:

```{r}
#| label: peak-lags
xc[, .SD[which.max(correlation)], by = component]
```

The peak correlation occurs at lag
`r xc[component == "expectation", lag[which.max(correlation)]]` for expectations,
`r xc[component == "climate", lag[which.max(correlation)]]` for the climate index, and
`r xc[component == "situation", lag[which.max(correlation)]]` for the situation assessment.
Because negative lags indicate a lead over production, only expectations provide a clear,
though short, lead. A formal test can determine whether this predictive direction holds after
accounting for the history of production itself.

### Criterion 2: Granger causality

A leading indicator should carry information about the *future* of the reference series.
The [Granger causality](https://en.wikipedia.org/wiki/Granger_causality) test asks whether past
values of the ifo index improve a forecast of production growth beyond production growth's own
past, and whether the reverse also holds. This tests predictive information, not an economic
causal mechanism.

```{r}
#| label: granger
pval <- function(x) format(signif(x[["Pr(>F)"]][2L], 2L), scientific = FALSE)

forward <- lmtest::grangertest(ip_growth ~ climate, order = 3L, data = activity)
reverse <- lmtest::grangertest(climate ~ ip_growth, order = 3L, data = activity)
forward
reverse
```

The predictive relationship is one-directional. Lags of the ifo climate improve the production
model (p = `r pval(forward)`), while lags of production do not improve the climate model
(p = `r pval(reverse)`).

Consistent with the cross-correlation, the forward-looking expectations component gives a
slightly stronger signal than the headline climate index.

```{r}
#| label: granger-expectation
lmtest::grangertest(ip_growth ~ expectation, order = 3L, data = activity)
```

### Criterion 3: forecast value

Does the ifo index add anything beyond what production growth already says about itself?
We compare an autoregressive benchmark that predicts production growth from its own three lags
against the same model augmented with three lags of the ifo climate index (the specification
underlying the Granger test above).
The comparison is in-sample: it measures fit, not out-of-sample forecast accuracy.

```{r}
#| label: forecast
dt <- activity |>
  _[, .(
    ip_growth,
    g1 = shift(ip_growth, 1L),
    g2 = shift(ip_growth, 2L),
    g3 = shift(ip_growth, 3L),
    c1 = shift(climate, 1L),
    c2 = shift(climate, 2L),
    c3 = shift(climate, 3L)
  )] |>
  na.omit()

fit_ar <- lm(ip_growth ~ g1 + g2 + g3, data = dt)
fit_ar_ifo <- lm(ip_growth ~ g1 + g2 + g3 + c1 + c2 + c3, data = dt)

rmse <- function(x) sqrt(mean(residuals(x)^2))
data.table(
  model = c("AR(3)", "AR(3) + ifo"),
  adj_r2 = c(summary(fit_ar)$adj.r.squared, summary(fit_ar_ifo)$adj.r.squared),
  rmse_in_sample = c(rmse(fit_ar), rmse(fit_ar_ifo))
)
```

Adding the climate index raises adjusted R² from
`r sprintf("%.3f", summary(fit_ar)$adj.r.squared)` to
`r sprintf("%.3f", summary(fit_ar_ifo)$adj.r.squared)` and lowers in-sample RMSE from
`r sprintf("%.2f", rmse(fit_ar))` to `r sprintf("%.2f", rmse(fit_ar_ifo))`.
The nested-model F test rejects the three climate coefficients being jointly zero
(p = `r pval(forward)`). This is the same test used in the forward Granger comparison above:

```{r}
#| label: forecast-test
anova(fit_ar, fit_ar_ifo)
```

### An out-of-sample check

In-sample fit does not guarantee a better forecast. We compare the Germany-wide climate,
situation, and expectations balances under the same expanding-window design. We also include
manufacturing expectations to test whether a more closely matched survey series improves the
forecast, as suggested by Lehmann's review. Starting with ten years of data, we re-estimate each
model monthly and predict the next observation.

```{r}
#| label: rolling-forecast
germany_components <- survey_long |>
  _[series == "balance"] |>
  dcast(yearmonth ~ indicator, value.var = "value") |>
  setnames(
    c("climate", "situation", "expectation"),
    c("de_climate", "de_situation", "de_expectation")
  )
manufacturing_expectation <- setDT(ifo_business("sectors"))[
  sector == "manufacturing" & series == "balance" & indicator == "expectation",
  .(yearmonth, mfg_expectation = value)
]
forecast_data <- mergelist(
  list(
    activity[, .(yearmonth, ip_growth)],
    germany_components,
    manufacturing_expectation
  ),
  on = "yearmonth",
  how = "inner"
)

lag_variables <- setdiff(names(forecast_data), "yearmonth")
lags <- 1:3
lag_columns <- paste0(rep(lag_variables, each = length(lags)), lags)
forecast_data[, (lag_columns) := shift(.SD, lags), .SDcols = lag_variables]

models <- list(
  "AR(3)" = paste0("ip_growth", lags),
  "Germany climate" = c(paste0("ip_growth", lags), paste0("de_climate", lags)),
  "Germany situation" = c(paste0("ip_growth", lags), paste0("de_situation", lags)),
  "Germany expectations" = c(paste0("ip_growth", lags), paste0("de_expectation", lags)),
  "Manufacturing expectations" = c(paste0("ip_growth", lags), paste0("mfg_expectation", lags))
)

required <- c("ip_growth", unique(unlist(models)))
forecast_data <- na.omit(forecast_data, cols = required)
initial_window <- 120L
forecast_months <- initial_window + seq_len(nrow(forecast_data) - initial_window)

predictions <- rbindlist(Map(
  function(predictors, model) {
    formula <- reformulate(predictors, response = "ip_growth")
    predicted <- vapply(
      forecast_months,
      function(month) {
        fit <- lm(formula, data = forecast_data[seq_len(month - 1L)])
        predict(fit, newdata = forecast_data[month])
      },
      numeric(1L)
    )
    data.table(
      model,
      yearmonth = forecast_data$yearmonth[forecast_months],
      actual = forecast_data$ip_growth[forecast_months],
      predicted
    )
  },
  predictors = models,
  model = names(models)
))

forecast_scores <- predictions |>
  _[, .(months = .N, rmse = sqrt(mean((actual - predicted)^2))), by = model] |>
  _[, improvement_over_ar := 100 * (rmse[model == "AR(3)"] - rmse) / rmse[model == "AR(3)"]] |>
  _[, let(rmse = round(rmse, 2L), improvement_over_ar = round(improvement_over_ar, 1L))]
forecast_scores[]
```

Across `r forecast_scores[model == "AR(3)", months]` forecasts, Germany-wide expectations lower
RMSE by `r forecast_scores[model == "Germany expectations", improvement_over_ar]`% relative to
the AR benchmark, slightly more than the climate balance at
`r forecast_scores[model == "Germany climate", improvement_over_ar]`%. The situation balance
raises RMSE by `r abs(forecast_scores[model == "Germany situation", improvement_over_ar])`%.
Manufacturing expectations lower it by
`r forecast_scores[model == "Manufacturing expectations", improvement_over_ar]`%.
The forecast ranking therefore agrees with the lead-lag analysis, although the differences
between the Germany-wide components are small. The closer sectoral match appears to matter, but
industrial production is broader than manufacturing alone. This remains a pseudo-out-of-sample
exercise because it uses the latest data vintage.

### When do the survey indicators help?

The average RMSE hides when the forecast gains occur. We accumulate the monthly difference in
squared errors between the AR benchmark and each model with expectations. The line rises when
the survey model produces the smaller error and falls when the benchmark does.

```{r}
#| label: forecast-stability
forecast_gain <- predictions |>
  _[, .(yearmonth, model, error = (actual - predicted)^2)] |>
  _[, benchmark_error := error[model == "AR(3)"], by = yearmonth] |>
  _[model %in% c("Germany expectations", "Manufacturing expectations")] |>
  setorder(model, yearmonth) |>
  _[, .(yearmonth, cumulative_gain = cumsum(benchmark_error - error)), by = model]

ggplot(forecast_gain, aes(x = yearmonth, y = cumulative_gain, color = model)) +
  annotate(
    "rect",
    xmin = as.Date("2020-01-01"),
    xmax = as.Date("2021-12-31"),
    ymin = -Inf,
    ymax = Inf,
    fill = "grey90",
    color = NA
  ) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_line() +
  labs(
    title = "Forecast gains are concentrated in the pandemic",
    subtitle = "Cumulative reduction in squared error relative to AR(3)",
    y = "Cumulative squared-error reduction",
    caption = sprintf(
      "Shaded area: January 2020 to December 2021.\nSources: ifo Business Survey and Deutsche Bundesbank, %s.",
      format(max(activity$yearmonth), "%B %Y")
    )
  ) +
  scale_color_manual(
    values = c("Germany expectations" = "darkred", "Manufacturing expectations" = "darkblue")
  ) +
  theme_ifo() +
  theme(axis.title.y = element_text())
```

### Caveats

**The pandemic has an outsized influence on the sample.** Most of the cumulative forecast gain
appears in 2020 and 2021, and both models lose some of that advantage after 2022. The full-sample
RMSE therefore does not show a stable forecasting advantage. A pre-2020 estimation or pandemic
indicators would provide a stricter sensitivity check.

**These results use the *final vintage* of industrial production.** The ifo index may be more
useful in real time because official production figures appear several weeks after the reference
month and are revised later. A real-time evaluation may therefore differ from the
pseudo-out-of-sample results above. See
[Lehmann and Reif (2021)](https://link.springer.com/article/10.1007/s41549-021-00055-5) for a
real-time comparison. This exercise demonstrates how to assess a leading indicator. It does not
settle the ifo survey's forecasting record.
